# Data Preprocessing Tutorial with R: Energy Load Analysis

In this tutorial, we'll walk through the preprocessing of energy load data using R. We will cover data loading, cleaning, feature engineering, and preparation for analysis, with explanations of each function used and exercises to reinforce learning.

------------------------------------------------------------------------

## 1. Loading Necessary Libraries

**Explanation:**

Before we begin, we need to load the libraries that provide the functions for data manipulation, reading files, handling dates, and performing rolling calculations.

-   **`dplyr`**: A grammar of data manipulation, providing a consistent set of verbs to solve data manipulation challenges.
-   **`readr`**: Provides a fast and friendly way to read rectangular data (like CSV files).
-   **`lubridate`**: Makes it easier to work with dates and times by providing functions that simplify date-time manipulation.
-   **`slider`**: Offers a family of functions to perform sliding (rolling) window operations.

```{r}
# Load necessary libraries
library(dplyr)
library(readr)
library(lubridate)
library(slider)
```

**Exercise 1:**

-   **Task**: Verify that you have the latest versions of these packages installed.
-   **Instructions**: Run `packageVersion("package_name")` for each package and compare it to the latest version on CRAN.
-   **Example**:

```{r}
packageVersion("dplyr")
packageVersion("readr")
packageVersion("lubridate")
packageVersion("slider")
```

------------------------------------------------------------------------

## 2. Loading the Data

**Explanation:**

We will load the energy load data from a CSV file available online. The data contains time-series information about energy consumption.

```{r}
# Load the data
data <- read.csv('https://raw.githubusercontent.com/HamidBekamiri/Econometrics-II/refs/heads/main/Data/time_series_60min_singleindex_filtered.csv')
```

**Functions Used:**

-   **`read.csv()`**: Reads a comma-separated values (CSV) file and creates a data frame (tibble) in R.

**Exercise 2:**

-   **Task**: Explore the structure and summary of the loaded data.
-   **Instructions**:
    -   Use `str(data)` to view the structure.
    -   Use `summary(data)` to get summary statistics.

```{r}
# Examine the structure of the data
str(data)

# Get summary statistics
summary(data)
```

------------------------------------------------------------------------

## 3. Preprocessing the Data

**Explanation:**

We need to convert the date-time information from strings to date objects and extract useful components like year, month, day, and hour. This will help us analyze the data based on these time units.

```{r}
# Preprocess the data
df_energy <- data %>%
  mutate(
    date = ymd_hms(utc_timestamp, tz = "UTC"),
    year = year(date),
    month = factor(month(date), levels = 1:12, labels = month.name),
    day_in_week = factor(wday(date, label = TRUE)),
    hour = factor(hour(date)),
    day_in_year = factor(yday(date)),
    week_in_year = factor(isoweek(date))
  ) %>%
  arrange(date) %>%
  drop_na()
```

**Functions Used:**

-   **`mutate()`**: Adds new variables or transforms existing ones.
-   **`ymd_hms()`**: Parses date-time strings into POSIXct date-time objects.
-   **`year()`, `month()`, `wday()`, `hour()`, `yday()`, `isoweek()`**: Extract specific components from date-time objects.
-   **`factor()`**: Converts variables into factors (categorical variables).
-   **`arrange()`**: Orders the rows by specified columns.
-   **`drop_na()`**: Removes rows containing missing values.

**Exercise 3:**

-   **Task**: Add a new column to `df_energy` indicating the quarter of the year.
-   **Instructions**:
    -   Use `quarter(date)` from `lubridate` to extract the quarter.
    -   Add it as a factor variable.

```{r}
df_energy <- df_energy %>%
  mutate(
    quarter = factor(quarter(date))
  )
```

------------------------------------------------------------------------

## 4. Creating Lag Features and Rolling Averages

**Explanation:**

Lag features help us include past values as predictors in our models. Rolling averages smooth out short-term fluctuations and highlight longer-term trends.

```{r}
# Create lag features and rolling averages
df_energy <- df_energy %>%
  mutate(
    lag_1 = lag(DE_load_actual_entsoe_transparency, 1),
    lag_2 = lag(DE_load_actual_entsoe_transparency, 2),
    rolling_avg_4 = slide_dbl(
      DE_load_actual_entsoe_transparency,
      mean,
      .before = 3,
      .complete = TRUE
    ),
    rolling_diff_4 = DE_load_actual_entsoe_transparency - rolling_avg_4
  ) %>%
  filter(
    !is.na(lag_1),
    !is.na(lag_2),
    !is.na(rolling_avg_4)
  )
```

**Functions Used:**

-   **`lag()`**: Shifts a vector by a specified number of positions.
-   **`slide_dbl()`**: Applies a function over a sliding window, returning a double vector.
-   **`filter()`**: Subsets rows based on specified conditions.

**Exercise 4:**

-   **Task**: Create lag features for 3 and 4 time steps back.
-   **Instructions**:
    -   Use `lag()` to create `lag_3` and `lag_4`.
    -   Add these to your `mutate()` function.

```{r}
df_energy <- df_energy %>%
  mutate(
    lag_3 = lag(DE_load_actual_entsoe_transparency, 3),
    lag_4 = lag(DE_load_actual_entsoe_transparency, 4)
  )
```

------------------------------------------------------------------------

## 5. Adding Holiday Variable for Weekends

**Explanation:**

Energy consumption patterns often differ on weekends and holidays. We'll create a binary variable to indicate weekends.

```{r}
# Add holiday variable for weekends
df_energy <- df_energy %>%
  mutate(
    holiday = if_else(day_in_week %in% c("Sat", "Sun"), 1, 0)
  )
```

**Functions Used:**

-   **`if_else()`**: A vectorized conditional function that returns a value depending on a logical condition.

**Exercise 5:**

-   **Task**: Create a new variable called `is_weekend` that is `TRUE` for Saturday and Sunday, and `FALSE` otherwise.
-   **Instructions**:
    -   Use `day_in_week` in combination with `%in%` and `mutate()`.

```{r}
df_energy <- df_energy %>%
  mutate(
    is_weekend = day_in_week %in% c("Sat", "Sun")
  )
```

------------------------------------------------------------------------

## 6. Selecting Relevant Columns

**Explanation:**

To focus on the most important variables for analysis, we'll select specific columns and ensure there are no missing values.

```{r}
# Select relevant columns
df_energy <- df_energy %>%
  select(
    year,
    month,
    day_in_week,
    hour,
    week_in_year,
    DE_load_actual_entsoe_transparency,
    lag_1,
    lag_2,
    rolling_avg_4,
    rolling_diff_4,
    holiday,
    date
  ) %>%
  drop_na()
```

**Functions Used:**

-   **`select()`**: Selects variables based on their names.

**Exercise 6:**

-   **Task**: Reorder the columns so that `date` is the first column in the data frame.
-   **Instructions**:
    -   Adjust the `select()` function to place `date` first.

```{r}
df_energy <- df_energy %>%
  select(
    date,
    year,
    month,
    day_in_week,
    hour,
    week_in_year,
    DE_load_actual_entsoe_transparency,
    lag_1,
    lag_2,
    rolling_avg_4,
    rolling_diff_4,
    holiday
  ) %>%
  drop_na()
```

------------------------------------------------------------------------

## 7. Viewing the First Few Rows and Subsetting

**Explanation:**

It's good practice to inspect your data after preprocessing. We'll look at the first few rows and reduce the data size for ease of handling.

```{r}
# View the first few rows
head(df_energy)
```

```{r}
# Subset to the first 5000 rows
df_energy <- df_energy[1:5000, ]

# Check the dimensions
dim(df_energy)
```

**Functions Used:**

-   **`head()`**: Displays the first few rows of a data frame.
-   **Subsetting**: Using `[rows, columns]` to select parts of a data frame.
-   **`dim()`**: Returns the dimensions (number of rows and columns) of the data frame.

**Exercise 7:**

-   **Task**: Find out how many observations and variables are in `df_energy` after subsetting.
-   **Instructions**:
    -   Use `nrow(df_energy)` and `ncol(df_energy)`.

```{r}
# Number of observations
nrow(df_energy)

# Number of variables
ncol(df_energy)
```

------------------------------------------------------------------------

**Additional Notes:**

-   **Data Types**: Be aware of the data types of your variables. Functions like `factor()` and `if_else()` are sensitive to data types.
-   **Date-Time Handling**: When working with time series data, proper date-time conversion is crucial for accurate analysis.
-   **Rolling Calculations**: Adjust the window size in rolling calculations to suit your analytical needs.

------------------------------------------------------------------------

## Summary

In this tutorial, we've covered:

1.  Loading essential R libraries.
2.  Importing data from a CSV file.
3.  Preprocessing data, including date-time conversion and feature extraction.
4.  Creating lag features and rolling averages.
5.  Adding categorical variables based on conditions.
6.  Selecting and rearranging relevant data columns.
7.  Inspecting and subsetting the data for analysis.

By completing the exercises, you've applied these steps and reinforced your understanding of data preprocessing in R.
